---
title: "SWELTR Depth 16S rRNA"
subtitle: "Scripts for filtering LotuS3 ASV data "
format:
  html: 
    code-copy: true
    code-overflow: wrap
    code-link: false
    css: styles.css
    embed-resources: true
    code-tools:
      source: true
      toggle: false
      caption: none
      
    theme: cosmo
---

```{r}
#| message: false
#| results: hide
#| code-fold: true
#| code-summary: "Click here for page build libraries and setup information."
#| label: setup
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
set.seed(919191)
library(PERFect)
library(microeco)
library(mia)
library(phyloseq)
library(microbiome)
pacman::p_load(magrittr, reactable, reactablefmtr, 
               janitor, tibble, readr,
               install = FALSE, update = FALSE)

options(scipen = 999)
knitr::opts_current$get(c(
  "cache",
  "cache.path",
  "cache.rebuild",
  "dependson",
  "autodep"
))
#root <- find_root(has_file("_quarto.yml"))
#source(file.path(root, "assets", "functions.R"))
objects()
```


```{r}
#| echo: false
#| eval: true
rm(list = ls())
load("16S/LOTUS3/lotus3_processing.rdata")
```













```{r}
#| echo: true
#| eval: false
load("16S/LOTUS3/phyloseq.Rdata")
objects()
physeq
me_asv <- readRDS("16S/LOTUS3/me_lotus3_asv.rds")
objects()

```


```{r}
data.frame(t(me_asv$otu_table))
```

```{r}
tmp_tab <- data.frame(me_asv$otu_table)

tmp_tab <- tmp_tab %>% tibble::rownames_to_column("ID")
#tmp_tab <- jamba::mixedSortDF(tmp_tab, decreasing = FALSE, 
#                                        useRownames = FALSE, byCols = 1)
tmp_tab <- tmp_tab %>% tibble::remove_rownames() 
tmp_tab <- tmp_tab %>% tibble::column_to_rownames("ID")
tmp_tab <- data.frame(t(tmp_tab))

tmp_tab_ord <- data.frame(me_asv$otu_table)
tmp_tab_ord <- tmp_tab_ord %>% tibble::rownames_to_column("ID")
tmp_tab_ord <- jamba::mixedSortDF(tmp_tab_ord, decreasing = TRUE, 
                                        useRownames = FALSE, byCols = 1)
tmp_tab_ord <- tmp_tab_ord %>% tibble::remove_rownames() 
tmp_tab_ord <- tmp_tab_ord %>% tibble::column_to_rownames("ID")
tmp_tab_ord <- data.frame(t(tmp_tab_ord))

tmp_tab_ord_asc <- data.frame(me_asv$otu_table)
tmp_tab_ord_asc <- tmp_tab_ord_asc %>% tibble::rownames_to_column("ID")
tmp_tab_ord_asc <- jamba::mixedSortDF(tmp_tab_ord_asc, decreasing = FALSE, 
                                        useRownames = FALSE, byCols = 1)
tmp_tab_ord_asc <- tmp_tab_ord_asc %>% tibble::remove_rownames() 
tmp_tab_ord_asc <- tmp_tab_ord_asc %>% tibble::column_to_rownames("ID")
tmp_tab_ord_asc <- data.frame(t(tmp_tab_ord_asc))

tmp_tab
tmp_tab_ord
tmp_tab_ord_asc
#identical(colnames(tmp_get_tab), colnames(tmp_get_tab_ord))

```

```{r}
ssu_per_pval <- 0.05
```

```{r}
tmp_sim <- PERFect_sim(X = tmp_tab, alpha = ssu_per_pval, Order = "NP", center = FALSE)
tmp_sim_ord <- PERFect_sim(X = tmp_tab_ord, alpha = ssu_per_pval, Order = "NP", center = FALSE)
tmp_sim_ord_asc <- PERFect_sim(X = tmp_tab_ord_asc, alpha = ssu_per_pval, Order = "NP", center = FALSE)

dim(tmp_sim$filtX)
dim(tmp_sim_ord$filtX)
dim(tmp_sim_ord_asc$filtX)



tmp_sim_pvals <- data.frame(tmp_sim$pvals) %>%
  tibble::rownames_to_column(var = "ASV_ID")
tmp_sim_ord_pvals <- data.frame(tmp_sim_ord$pvals) %>%
  tibble::rownames_to_column(var = "ASV_ID")
tmp_sim_ord_asc_pvals <- data.frame(tmp_sim_ord_asc$pvals) %>%
  tibble::rownames_to_column(var = "ASV_ID")


tmp_pvals <- dplyr::full_join(tmp_sim_ord_asc_pvals, tmp_sim_ord_pvals, by = "ASV_ID") %>%
  dplyr::full_join(., tmp_sim_pvals, by = "ASV_ID")
write.table(tmp_pvals, "tmp_pvals.txt", sep = "\t", row.names = FALSE, quote = FALSE)
```



```{r}

tmp_tab
tmp_tab_ord
tmp_tab_ord_asc
tmp_tab_lc <- phyloseq::genefilter_sample(
                                         as.matrix(t(tmp_tab)), 
                                         filterfun_sample(function(x) x >= 5), 
                                         A = 0.2 * 39)




tmp_arb_bool <- data.frame(tmp_tab_lc) %>%
  tibble::rownames_to_column(var = "ASV_ID")


tmp_tab_lc_true <- data.frame(tmp_tab_lc) %>%
  dplyr::filter(data.frame(tmp_tab_lc) == "TRUE")


tmp_pvals <- dplyr::full_join(tmp_sim_ord_asc_pvals, tmp_sim_ord_pvals, by = "ASV_ID") %>%
  dplyr::full_join(., tmp_sim_pvals, by = "ASV_ID") %>%
  dplyr::full_join(., tmp_arb_bool, by = "ASV_ID")


write.table(tmp_pvals, "tmp_pvals.txt", sep = "\t", row.names = FALSE, quote = FALSE)


tmp_low_count <- phyloseq::prune_taxa(tmp_low_count, ssu18_ps_work)
tmp_low_var <- phyloseq::filter_taxa(tmp_low_count, 
                                     function(x) var(x) > 0.2, prune = TRUE)
ssu18_ps_filt <- tmp_low_var
ssu18_ps_filt@phy_tree <- NULL
tmp_tree <- rtree(ntaxa(ssu18_ps_filt), rooted = TRUE,
                      tip.label = taxa_names(ssu18_ps_filt))
ssu18_ps_filt <- merge_phyloseq(ssu18_ps_filt,
                          sample_data,
                          tmp_tree)

rm(list = ls(pattern = "tmp_"))
ssu18_ps_filt
```






























```{r}
tmp_df <- data.frame(t(otu_table(physeq)))

tmp_counts <- mock2$Counts %>% as.matrix() %>% as.table()
res_sim <- PERFect_sim(Counts)

tmp_df1 <- adorn_totals(tmp_df, where = "col", name = "Sum")
tmp_df1 <- as.data.frame(tibble::rownames_to_column(tmp_df1))
new_df <- data.frame(tmp_df1[, c(1, ncol(tmp_df1))])
write.table(new_df, "16S/LOTUS3/lotus3_otu.txt", row.names = FALSE,  sep = "\t", quote = FALSE)

tmp_df1 <- readRDS("16S/DADA2/seqtab.nochim_pseudo.rds")
tmp_df1 <- as.data.frame(tmp_df1)
tmp_df1 <- adorn_totals(tmp_df1, where = "col", name = "Sum")
tmp_df1 <- as.data.frame(tibble::rownames_to_column(tmp_df1))
new_df <- data.frame(tmp_df1[, c(1, ncol(tmp_df1))])
write.table(new_df, 16S/DADA2/"dada_otu.txt", row.names = FALSE,  sep = "\t", quote = FALSE)
```

```{r}
data.frame(tax_table(physeq))
physeq_tab <- data.frame(otu_table(physeq))
physeq_tab <- physeq_tab %>% tibble::rownames_to_column("ID")
#physeq_tab <- jamba::mixedSortDF(physeq_tab, decreasing = FALSE, 
#                                      useRownames = FALSE, byCols = 1)
physeq_tab <- physeq_tab %>% tibble::remove_rownames() 
physeq_tab <- physeq_tab %>% tibble::column_to_rownames("ID")
physeq_tab <- data.frame(t(physeq_tab))
```

```{r}
physeq_tab_ord <- data.frame(otu_table(physeq))
physeq_tab_ord <- physeq_tab_ord %>% tibble::rownames_to_column("ID")
physeq_tab_ord <- jamba::mixedSortDF(physeq_tab_ord, decreasing = FALSE, 
                                      useRownames = FALSE, byCols = 1)
physeq_tab_ord <- physeq_tab_ord %>% tibble::remove_rownames() 
physeq_tab_ord <- physeq_tab_ord %>% tibble::column_to_rownames("ID")
physeq_tab_ord <- data.frame(t(physeq_tab_ord))
```

```{r}
tmp_tab_name <- purrr::map_chr(i, ~ paste0(., "_perfect"))
assign(tmp_tab_name, physeq_tab)

tmp_tab_ord_name <- purrr::map_chr(i, ~ paste0(., "_ord_perfect"))
assign(tmp_tab_ord_name, physeq_tab_ord)

objects()
```
## Required Packages





```
phyloseq-class experiment-level object
otu_table()   OTU Table:         [ 12091 taxa and 41 samples ]
sample_data() Sample Data:       [ 41 samples by 14 sample variables ]
tax_table()   Taxonomy Table:    [ 12091 taxa by 7 taxonomic ranks ]
phy_tree()    Phylogenetic Tree: [ 12091 tips and 12089 internal nodes ]
```

```{r}
#| echo: false
#| eval: false
min_read_ps <- min(readcount(physeq))
max_read_ps <- max(readcount(physeq))
total_reads_ps <- sum(readcount(physeq))
mean_reads_ps <- round(mean(readcount(physeq)), digits = 0)
median_reads_ps <- median(readcount(physeq))
total_asvs_ps <- ntaxa(physeq)
singleton_ps <- tryCatch(ntaxa(rare(physeq, detection = 1, 
                                    prevalence = 0)), 
                         error=function(err) NA)
singleton_ps_perc <- tryCatch(round((100*(ntaxa(rare(physeq, detection = 1, prevalence = 0)) /
                                   ntaxa(physeq))), digits = 3), error=function(err) NA)
sparsity_ps <- round(length(which(abundances(physeq) == 0))/length(abundances(physeq)),
                     digits = 3)

ds_metrics <- data.frame(rbind(min_read_ps, max_read_ps, total_reads_ps, 
                               mean_reads_ps, median_reads_ps, total_asvs_ps, 
                               singleton_ps, singleton_ps_perc, sparsity_ps))
```

| Metric                              | Results                             |
|-------------------------------------|-------------------------------------|
| Min. number of reads                | `r ds_metrics[1, ]`                 |
| Max. number of reads                | `r ds_metrics[2, ]`                 |
| Total number of reads               | `r ds_metrics[3, ]`                 |
| Average number of reads             | `r ds_metrics[4, ]`                 |
| Median number of reads              | `r ds_metrics[5, ]`                 |
| Total ASVs                          | `r ds_metrics[6, ]`                 |
| Number of singleton ASVs            | `r ds_metrics[7, ]`                 |
| Any ASVs sum to 1 or less?          | `r isTRUE(ds_metrics[7, ] >= 1)`    |
| Percent of ASVs that are singletons | `r ds_metrics[8, ]`                 |
| Sparsity                            | `r ds_metrics[9, ]`                 |

In this next part of the workflow our main goal is to create a *microtable object* using the R package [microeco](https://chiliubio.github.io/microeco_tutorial/) [@liu2021microeco]. The microtable will be used to store the ASV by sample data as well the taxonomic, fasta, and sample data in a single object. More on that in a moment.

We will also:

-   Remove any ASVs without kingdom level classification.\
-   Revome any contaminants (chloroplast, mitochondria, etc.).\
-   Remove Negative Control (NC) samples.\
-   Remove any low-count samples.

## Sample Summary

Before we begin, let's create a summary table containing some basic sample metadata and the read count data. We want to inspect how total reads changed through the workflow. Table headers are as follows:

| Header         | Description                                             |
|----------------|---------------------------------------------------------|
| `SampleID`     | New sample ID based on plot, depth, treatment, & pair   |
| `Plot`         | Experimental plot ID                                    |
| `Treatment`    | Control vs warming                                      |
| `Temp`         | Warming treatment temperature                           |
| `Depth`        | Sampling depth in soil                                  |
| `Pairing`      | Sample pairings                                         |
| `raw_rc`       | No. of raw reads                                        |
| `cutadapt_rc`  | Read count after cutdapt                                |
| `final_rc`     | Read count after LotuS3 pipeline                        |
| `no_asv`       | No. of ASVs                                             |

<br/>

Sample data 

```{r}
#| echo: true
tmp_otu <- data.frame(otu_table(physeq))
tmp_otu_sum <- data.frame(colSums(tmp_otu > 0)) %>%
  tibble::rownames_to_column("SampleID")
names(tmp_otu_sum)[2] <- "no_asv"
tmp_read_sum <- data.frame(colSums(tmp_otu)) %>%
  tibble::rownames_to_column("SampleID")
names(tmp_read_sum)[2] <- "final_rc"
tmp_track_reads <- read.table(
    "16S/CUTADAPT/ssu_cutadapt_track.txt",
    header = TRUE, sep = "\t"
)

tmp_track_reads$SampleID <- stringr::str_replace_all(
  tmp_track_reads$SampleID, "-", "_")
names(tmp_track_reads)[2] <- "raw_rc"
names(tmp_track_reads)[3] <- "cutadapt_rc"
#----------------/-------------------------#
tmp_samp <- data.frame(sample_data(physeq)) %>% 
  tibble::rownames_to_column("SampleID")
#----------------/-------------------------#
sample_tab <- dplyr::left_join(tmp_samp, tmp_track_reads, by = "SampleID") %>% 
  dplyr::left_join(., tmp_read_sum, by = "SampleID") %>% 
  dplyr::left_join(., tmp_otu_sum, by = "SampleID") 

sample_tab <- sample_tab[, -c(2, 3, 9:15)]

sample_tab$per_reads_kept <- round(sample_tab$final_rc/sample_tab$raw_rc, digits = 3)
sample_tab <- sample_tab %>% dplyr::relocate(per_reads_kept, .before = no_asv)

#rm(list = ls(pattern = "tmp_"))
```

```{r}
#| echo: false
#| eval: true
reactable(
    sample_tab, defaultColDef = colDef(
        header = function(value) gsub("_", " ", value, fixed = TRUE),
        cell = function(value) format(value, nsmall = 0),
        align = "center", filterable = TRUE, sortable = TRUE,
        resizable = TRUE, footerStyle = list(fontWeight = "bold")
    ),
    columns = list(
        SampleID = colDef(
            name = "SampleID", sticky = "left", style = list(borderRight = "1px solid #eee"),
            headerStyle = list(borderRight = "1px solid #eee"),
            align = "left", minWidth = 150, footer = "Total reads"),
        raw_rc = colDef(name = "raw rc", footer = function(values) sprintf("%.0f", sum(values))),
        cutadapt_rc = colDef(name = "cutadapt rc", footer = function(values) sprintf("%.0f", sum(values))),
        final_rc = colDef(name = "final rc", footer = function(values) sprintf("%.0f", sum(values))),
        per_reads_kept = colDef(name = "per reads retain"),
        no_asv = colDef(name = "total ASVs")
    ),
    searchable = TRUE, defaultPageSize = 5, pageSizeOptions = c(5, 10, nrow(sample_tab)),
    showPageSizeOptions = TRUE, highlight = TRUE, bordered = TRUE,
    striped = TRUE, compact = FALSE, wrap = FALSE, showSortable = TRUE,
    fullWidth = TRUE, theme = reactableTheme(style = list(fontSize = "0.8em"))
) %>%
    reactablefmtr::add_subtitle(
        "Sample metadata including read changes at start and end of workflow.",
        font_size = 15
    )
```

```{r}
#| echo: false
#| eval: false
readr::write_delim(sample_tab, "16S/LOTUS3/sample_data_with_rc.txt",
    delim = "\t")
```

</br>

{{< downloadthis 16S/LOTUS3/sample_data_with_rc.txt dname=sample_data_with_rc label="Download sample metadata" icon=table type=info class=data-button id=shrimp_md >}}

## Prep Data for `microeco`

Like any tool, the microeco package needs the data in a specific form. I formatted our data to match the mock data in the microeco tutorial, specifically [this section](https://chiliubio.github.io/microeco_tutorial/basic-class.html#prepare-the-example-data).

### A. Taxonomy Table

Here is what the taxonomy table looks like in the mock data.

```{r}
#| eval: false
#| echo: true
tmp_tax <- data.frame(tax_table(physeq))
tmp_tax[1:6, 1:4]
```

```         
           Domain   Phylum      Class      Order
ASV11944   Bacteria Bacillota   Bacilli    RES148
ASV8741    Bacteria Bacillota_A Clostridia Christensenellales
ASV11355   Bacteria Bacillota_A Clostridia Christensenellales
ASV6303    Bacteria Bacillota_A Clostridia Christensenellales
ASV9863    Bacteria Bacillota_A Clostridia Christensenellales
ASV11366   Bacteria Bacillota_A Clostridia Christensenellales
```

Next we need to add rank definitions to each classification.

```{r}
#| echo: true
#| eval: false
tmp_tax <- data.frame(tax_table(physeq))   
tmp_tax <- tmp_tax %>% dplyr::rename("Kingdom" = "Domain")
tmp_tax <- tmp_tax %>%
  dplyr::mutate_all(~stringr::str_replace_all(., "\\?", ""))

tmp_tax$Kingdom <- paste("k", tmp_tax$Kingdom, sep = "__")
tmp_tax$Phylum <- paste("p", tmp_tax$Phylum, sep = "__")
tmp_tax$Class <- paste("c", tmp_tax$Class, sep = "__")
tmp_tax$Order <- paste("o", tmp_tax$Order, sep = "__")
tmp_tax$Family <- paste("f", tmp_tax$Family, sep = "__")
tmp_tax$Genus <- paste("g", tmp_tax$Genus, sep = "__")
tmp_tax$Species <- paste("s", tmp_tax$Species, sep = "__")
```

And now the final, modified taxonomy table.

```{r}
#| echo: true
#| eval: false
tmp_tax[1:6, 1:4]
```

```
            Kingdom     Phylum          Class         Order
ASV11944	  k__Bacteria	p__Bacillota	  c__Bacilli	  o__RES148
ASV8741	    k__Bacteria	p__Bacillota_A	c__Clostridia	o__Christensenellales
ASV11355	  k__Bacteria	p__Bacillota_A	c__Clostridia	o__Christensenellales
ASV6303	    k__Bacteria	p__Bacillota_A	c__Clostridia	o__Christensenellales
ASV9863	    k__Bacteria	p__Bacillota_A	c__Clostridia	o__Christensenellales
ASV11366	  k__Bacteria	p__Bacillota_A	c__Clostridia	o__Christensenellales
```

### B. Sequence Table

Here is what the sequence table looks like in the mock data.

```{r}
#| eval: false
#| echo: true
tmp_otu[1:6, 1:3]
```

```
          P00_D00_000_NNN P01_D00_010_W4A P01_D10_020_W4A
ASV11944                0	            1	            0	
ASV8741	                0	            3            	1	
ASV11355                0	            2	            0	
ASV6303	                0	            5	            0	
ASV9863	                0	            8	            1	
ASV11366                0	            0	            0	
```

```{r}
#| echo: true
#| eval: false
identical(row.names(tmp_otu), row.names(tmp_tax))
```

```
[1] TRUE
```

### C. Sample Table

Here is what the sample table looks like.

```{r}
#| eval: true
#| echo: true
head(sample_tab)
```

```{r}
#| echo: true
#| eval: false
sample_tab <- sample_tab %>% tibble::column_to_rownames("SampleID")
sample_tab$SampleID <- rownames(sample_tab)
sample_tab <- sample_tab %>% dplyr::relocate(SampleID)
```

And now the final, modified sample table.

## Create a Microtable Object

With these three files in hand we are now ready to create a microtable object.

::: callout-note
A microtable object contains an ASV table (taxa abundances), sample metadata, and taxonomy table (mapping between ASVs and higher-level taxonomic classifications).
:::

```{r}
#| echo: true
#| eval: false
sample_info <- sample_tab
tax_tab <- tmp_tax
otu_tab <- tmp_otu
```

```{r}
#| echo: true
#| eval: false
library(microeco)
tmp_me <- microtable$new(sample_table = sample_info, 
                         otu_table = otu_tab, 
                         tax_table = tax_tab)

```

```         
microtable-class object:
sample_table have 41 rows and 11 columns
otu_table have 12091 rows and 41 columns
tax_table have 12091 rows and 7 columns
```

### Add Representative Sequence

We can also add representative sequences for each OTU/ASV. For this step, we can simply grab the sequences from the row names of the DADA2 taxonomy object loaded above.


```{r}
#| echo: true
#| eval: false
rep_fasta <- Biostrings::readDNAStringSet("16S/LOTUS3/OTU.fna")
identical(row.names(tmp_me$tax_table), row.names(tmp_me$otu_table))
```

```
[1] TRUE
```

```{r}
#| echo: true
#| eval: false
tmp_fasta_names <- names(rep_fasta)
tmp_rep_fasta <- rep_fasta[match(row.names(tmp_me$tax_table), tmp_fasta_names)]
rep_fasta <- tmp_rep_fasta
tmp_me$rep_fasta <- rep_fasta
identical(row.names(tmp_me$tax_table), names(rep_fasta))
tmp_me$tidy_dataset()
```

```
[1] TRUE
```

```{r}
#| echo: true
#| eval: false
me_asv_raw <- microeco::clone(tmp_me)
me_asv_raw
```

```{r}
#| echo: false
#| eval: true
me_asv_raw
```

> The microeco object `me_asv_raw` contains all data from the LotuS3 pipeline, before any dataset curation. 

## Curate the Data Set

Pretty much the last thing to do is remove unwanted taxa, negative controls, and low-count samples.

### Remove any Kingdom NAs

Here we can just use the straight up `subset` command since we do not need to worry about any ranks above Kingdom also being removed.

```{r}
#| echo: true
#| eval: false
tmp_no_na <- microeco::clone(tmp_me)
tmp_no_na$tax_table %<>% 
  base::subset(Kingdom == "k__Archaea" | Kingdom == "k__Bacteria")
tmp_no_na$tidy_dataset()
```

```{r}
#| echo: true
#| eval: false
me_asv_no_na <- microeco::clone(tmp_no_na)
me_asv_no_na
```

```{r}
#| echo: false
#| eval: true
me_asv_no_na
```

> The microeco object `me_asv_no_na` conatins only ASVs calssified as Archaea or Bacteria. 

### Remove Contaminants

Now we can remove any potential contaminants like mitochondria or chloroplasts.

```{r}
#| echo: true
#| eval: false
tmp_no_cont <- microeco::clone(tmp_no_na)
tmp_no_cont$filter_pollution(taxa = c("mitochondria", "chloroplast"))
tmp_no_cont$tidy_dataset()
```

```         
Total 17 features are removed from tax_table ...
```

```{r}
#| echo: true
#| eval: false
me_asv_no_cont <- microeco::clone(tmp_no_cont)
me_asv_no_cont
```

```{r}
#| echo: false
#| eval: true
me_asv_no_cont
```

> The microeco object `me_asv_no_cont` does not conatins contain ASVs calssified as mitochondria or chloroplast. 

### Remove Negative Controls (NC)

Now we need to remove the NC samples *and* ASVs found in those sample. We first identified all ASVs that were present in at least one NC sample represented by at least 1 read. We did this by subsetting the NC samples from the new microtable object.

```{r}
#| echo: true
#| eval: false
tmp_nc <- microeco::clone(tmp_no_cont)
tmp_nc$sample_table <- subset(tmp_nc$sample_table, Treatment == "Negative")
tmp_nc$tidy_dataset()
```

```
11974 taxa with 0 abundance are removed from the otu_table ...
```

```{r}
#| echo: true
#| eval: false
me_asv_nc <- microeco::clone(tmp_nc)
me_asv_nc
```

```{r}
#| echo: false
#| eval: true
me_asv_nc
```

> The microeco object `me_asv_nc` conatins only the negative control sample(s) and associated ASVs. 

Looks like there are `r nrow(me_asv_nc$tax_table)` ASVs in the NC samples from a total of `r sum(me_asv_nc$taxa_sums())` reads.

```{r}
#| echo: true
#| eval: false
nc_asvs <- row.names(tmp_nc$tax_table)
nc_asvs
```

```{r}
#| echo: false
#| eval: true
nc_asvs
```

ASVs are numbered in order by total abundance in the data set so we know that many of the ASVs in the NC samples are not particularly abundant in the dataset. We can look at the abundance of these ASVs across all samples and compare it to the NC. This takes a bit of wrangling.

Essentially, for each ASV, the code below calculates:

-   The total number of NC samples containing at least 1 read.\
-   The total number of reads in NC samples.\
-   The total number of non-NC samples containing at least 1 read.\
-   The total number of reads in non-NC samples.\
-   The percent of reads in the NC samples and the percent of NC samples containing reads.

```{r}
#| echo: true
#| eval: false
tmp_rem_nc <- microeco::clone(tmp_no_cont)
tmp_rem_nc_df <- tmp_rem_nc$otu_table
tmp_rem_nc_df <- tmp_rem_nc_df %>% dplyr::filter(row.names(tmp_rem_nc_df) %in% nc_asvs)
tmp_rem_nc_df <- tmp_rem_nc_df %>% tibble::rownames_to_column("ASV_ID")
```

```{r}
#| echo: true
#| eval: false
tmp_rem_nc_df <- tmp_rem_nc_df  %>% 
  dplyr::mutate(total_reads_NC = rowSums(dplyr::select(., contains("NNN"))), 
         .after = "ASV_ID")
tmp_rem_nc_df <- dplyr::select(tmp_rem_nc_df, -contains("NNN"))
tmp_rem_nc_df <- tmp_rem_nc_df %>%
  dplyr::mutate(total_reads_samps = rowSums(.[3:ncol(tmp_rem_nc_df)]), 
                .after = "total_reads_NC")
tmp_rem_nc_df[, 4:ncol(tmp_rem_nc_df)] <- list(NULL)
tmp_rem_nc_df <- tmp_rem_nc_df %>%
  dplyr::mutate(perc_in_neg = 100*(
    total_reads_NC / (total_reads_NC + total_reads_samps)),
                .after = "total_reads_samps")
```

```{r}
#| echo: true
#| eval: false
tmp_rem_nc_df$perc_in_neg <- round(tmp_rem_nc_df$perc_in_neg, digits = 6)

tmp_1 <- data.frame(rowSums(tmp_rem_nc$otu_table != 0))
tmp_1 <- tmp_1 %>% tibble::rownames_to_column("ASV_ID")
tmp_1 <- tmp_1 %>% dplyr::rename("total_samples" = 2)  

tmp_2 <- dplyr::select(tmp_rem_nc$otu_table, contains("NNN"))
tmp_2$num_samp_nc <- rowSums(tmp_2 != 0)
tmp_2 <- dplyr::select(tmp_2, contains("num_samp_nc"))
tmp_2 <- tmp_2 %>% tibble::rownames_to_column("ASV_ID")

tmp_3 <- dplyr::select(tmp_rem_nc$otu_table, -contains("NNN"))
tmp_3$num_samp_no_nc <- rowSums(tmp_3 != 0)
tmp_3 <- dplyr::select(tmp_3, contains("num_samp_no_nc"))
tmp_3 <- tmp_3 %>% tibble::rownames_to_column("ASV_ID")

tmp_rem_nc_df <- dplyr::left_join(tmp_rem_nc_df, tmp_1) %>%
                 dplyr::left_join(., tmp_2) %>%
                 dplyr::left_join(., tmp_3)

tmp_rem_nc_df <- tmp_rem_nc_df %>%
  dplyr::mutate(perc_in_neg_samp = 100*( num_samp_nc / (num_samp_nc + num_samp_no_nc)),
                .after = "num_samp_no_nc")
```

```{r}
#| echo: true
#| eval: false
nc_check <- tmp_rem_nc_df
```

```{r}
#| echo: false
#| eval: true
  reactable(nc_check,
  defaultColDef = colDef(
    header = function(value) gsub("_", " ", value, fixed = TRUE),
    #cell = function(value) format(value, nsmall = 0),
    align = "center", filterable = FALSE, sortable = TRUE, resizable = TRUE,
    footerStyle = list(fontWeight = "bold")
    ), 
  columns = list(
    ASV_ID = colDef(name = "ASV ID", 
                       sticky = "left", 
                       style = list(borderRight = "1px solid #eee"),
                       headerStyle = list(borderRight = "1px solid #eee"), 
                       align = "left",
                       minWidth = 100),
  total_reads_NC = colDef(name = "reads in NC"),
  total_reads_samps = colDef(name = "reads in non NC"),
  perc_in_neg = colDef(name = "% in NC", format = colFormat(digits = 4)),
  total_samples = colDef(name = "total samples"),
  num_samp_nc = colDef(name = "Total NC samples"),
  num_samp_no_nc = colDef(name = "Total non-NC samples"),
  perc_in_neg_samp = colDef(name = "% of samples", format = colFormat(digits = 4))
  ),
  searchable = FALSE, defaultPageSize = 11, 
  pageSizeOptions = c(5, 10, nrow(nc_check)), 
  showPageSizeOptions = TRUE, highlight = TRUE, 
  bordered = TRUE, striped = TRUE, compact = FALSE, 
  wrap = FALSE, showSortable = TRUE, fullWidth = TRUE,
  theme = reactableTheme(style = list(fontSize = "0.8em"))) %>%
  reactablefmtr::add_subtitle("Summary of ASVs detected in Negative Control (NC) samples.", 
                              font_size = 15)
```

```{r}
#| echo: false
#| eval: false
write_delim(nc_check, "16S/LOTUS3/asv_in_nc_samples.txt",
    delim = "\t")
```

{{< downloadthis 16S/LOTUS3/asv_in_nc_samples.txt dname=asv_in_nc_samples label="Download summary of ASVs detected in Negative Control (NC) samples" icon=table type=info class=data-button id=asv_in_nc_samples >}}

Looking at these data we can see that ASVs like ASV1 are only represented by a really small number of NC reads and samples. On the other hand, ASVs such as ASV6060, ASV9669, and ASV9992 are very abundant in NC samples. We decided to remove ASVs if:

-   The number of reads found in NC samples accounted for more than 10% of total reads OR
-   The percent of NC samples containing the ASV was greater than 10% of total samples.

```{r}
#| echo: true
#| eval: false
nc_remove <- nc_check %>% 
  dplyr::filter(perc_in_neg > 10 | perc_in_neg_samp > 10)
#tmp_rem_asv <- nc_remove$ASV_ID %>% 
#  unlist(strsplit(., split = ", ")) 
```

```{r}
#| echo: false
#| eval: false
nc_remain <- dplyr::anti_join(nc_check, nc_remove)

rem_nc_reads <- sum(nc_remove$total_reads_NC)
rem_sam_reads <- sum(nc_remove$total_reads_samps)
per_reads_rem <- round(100*( rem_nc_reads / (rem_nc_reads + rem_sam_reads)), 
                       digits = 3)

ret_nc_reads <- sum(nc_remain$total_reads_NC)
ret_sam_reads <- sum(nc_remain$total_reads_samps)
per_reads_ret <- round(100*( ret_nc_reads / (ret_nc_reads + ret_sam_reads)), 
                       digits = 3)
```

|          | Total ASVs          | NC reads         | non NC reads      | \% NC reads       |
|----------|---------------------|------------------|-------------------|-------------------|
| Removed  | `r nrow(nc_remove)` | `r rem_nc_reads` | `r rem_sam_reads` | `r per_reads_rem` |
| Retained | `r nrow(nc_remain)` | `r ret_nc_reads` | `r ret_sam_reads` | `r per_reads_ret` |

We identified a total of `r nrow(nc_check)` ASVs that were present in at least 1 NC sample by at least 1 read. We removed any ASV where more than 10% of total reads were found in NC samples OR any ASV found in more than 10% of NC samples. Based on these criteria we removed `r nrow(nc_remove)` ASVs from the data set, which represented `r rem_nc_reads` total reads in NC samples and `r rem_sam_reads` total reads in non-NC samples. Of the total reads removed `r per_reads_rem`% came from NC samples. Of all ASVs identified in NC samples,`r nrow(nc_remain)` were retained because the fell below the threshhold criteria. These ASVs accounted for `r ret_nc_reads` reads in NC samples and `r ret_sam_reads` reads in non-NC samples. NC samples accounted for `r per_reads_ret`% of these reads.

OK, now we can remove the NC samples and any ASVs that met our criteria described above.

```{r}
#| echo: true
#| eval: false
tmp_no_nc <- microeco::clone(tmp_no_cont)

tmp_rem_asv <- as.factor(nc_remove$ASV_ID)
tmp_no_nc$otu_table <- tmp_rem_nc$otu_table %>% 
  dplyr::filter(!row.names(tmp_no_nc$otu_table) %in% tmp_rem_asv)
tmp_no_nc$tidy_dataset()

tmp_no_nc$sample_table <- subset(tmp_no_nc$sample_table, 
                                 Treatment != "Negative")
tmp_no_nc$tidy_dataset()
```

```{r}
#| echo: true
#| eval: false
me_asv_no_nc <- microeco::clone(tmp_no_nc)
me_asv_no_nc
```

```{r}
#| echo: false
#| eval: true
me_asv_no_nc
```

> The microeco object `me_asv_no_nc` does not conatins the negative control sample(s). 

### Remove Low-Count Samples

Next, we can remove samples with really low read counts---here we set the threshold to `1000` reads.

```{r}
#| echo: true
#| eval: false
tmp_no_low <- microeco::clone(tmp_no_nc)
tmp_no_low$otu_table <- tmp_no_nc$otu_table %>%
          dplyr::select(where(~ is.numeric(.) && sum(.) >= 1000))
tmp_no_low$tidy_dataset()
```

Check if any row sum (i.e., ASVs) is equal to 0 after removing the negative control sample(s).

```{r}
tmp_row_sums <- rowSums(tmp_no_low$otu_table)
any(tmp_row_sums == 0)
```

```
[1] FALSE
```

```{r}
#| echo: true
#| eval: false
me_asv_no_low <- microeco::clone(tmp_no_low)
me_asv_no_low
```

```{r}
#| echo: false
#| eval: true
me_asv_no_low
```

> The microeco object `me_asv_no_low` does not conatins the low abundance samples. 

Giving us the final microtable object.

```{r}
#| echo: true
#| eval: false
me_asv <- microeco::clone(me_asv_no_low)
```

```{r}
#| echo: false
#| eval: true
me_asv
```

## Summary

Now time to summarize the data. For this we use the R package [miaverse](https://microbiome.github.io) [@felix2024mia].

```{r}
#| echo: false
#| eval: false
identical(rownames(me_asv_raw$sample_table), colnames(me_asv_raw$otu_table))
identical(rownames(me_asv$sample_table), colnames(me_asv$otu_table))
```

First we do a little formatting to get our data compatible with mia.

```{r}
#| echo: true
#| eval: false
# https://github.com/microbiome/OMA/issues/202
tmp_counts <- as.matrix(me_asv_raw$otu_table)
tmp_assays <-  SimpleList(counts = tmp_counts)
mia_me_asv_raw <- TreeSummarizedExperiment(assays = tmp_assays,
                                colData = DataFrame(me_asv_raw$sample_table),
                                rowData = DataFrame(me_asv_raw$tax_table))
rm(list = ls(pattern = "tmp_"))
tmp_counts <- as.matrix(me_asv$otu_table)
tmp_assays <-  SimpleList(counts = tmp_counts)
mia_me_asv <- TreeSummarizedExperiment(assays = tmp_assays,
                                colData = DataFrame(me_asv$sample_table),
                                rowData = DataFrame(me_asv$tax_table))
mia_me_asv_raw_summ <- summary(mia_me_asv_raw, assay.type = "counts")
mia_me_asv_summ <- summary(mia_me_asv, assay.type = "counts")
rm(list = ls(pattern = "tmp_"))
```

```{r}
#| echo: false
#| eval: false
me_dataset <- c("original", "No NA kingdom",
                "no contaminants", "no neg control", 
                "no low count samps"
                )

total_asvs <- c(
                nrow(me_asv_raw$tax_table), 
                nrow(me_asv_no_na$tax_table), 
                nrow(me_asv_no_cont$tax_table), 
                nrow(me_asv_no_nc$tax_table), 
                nrow(me_asv_no_low$tax_table)
                )

total_reads <- c(
                sum(me_asv_raw$sample_sums()), 
                sum(me_asv_no_na$sample_sums()), 
                sum(me_asv_no_cont$sample_sums()), 
                sum(me_asv_no_nc$sample_sums()),
                sum(me_asv_no_low$sample_sums())
                )

total_samples <- c(
                nrow(me_asv_raw$sample_table), 
                nrow(me_asv_no_na$sample_table), 
                nrow(me_asv_no_cont$sample_table), 
                nrow(me_asv_no_nc$sample_table), 
                nrow(me_asv_no_low$sample_table) 
                )
pipe_summary <- data.frame(cbind(me_dataset, total_asvs, 
                                 total_reads, total_samples
                                 )
                           )
```

```{r}
#| echo: false
#| eval: true
pipe_summary
```


| Metric                | Start                                      | End                                    |
|-----------------------|--------------------------------------------|----------------------------------------|
| Min. number reads     | `r mia_me_asv_raw_summ$samples[2]`         | `r mia_me_asv_summ$samples[2]`         |
| Max. number reads     | `r mia_me_asv_raw_summ$samples[3]`         | `r mia_me_asv_summ$samples[3]`         |
| Total number reads    | `r mia_me_asv_raw_summ$samples[1]`         | `r mia_me_asv_summ$samples[1]`         |
| Avg number reads      | `r round(mia_me_asv_raw_summ$samples[5])`  | `r round(mia_me_asv_summ$samples[5])`  |
| Median no. reads      | `r mia_me_asv_raw_summ$samples[4]`         | `r mia_me_asv_summ$samples[4]`         |
| Total ASVs            | `r mia_me_asv_raw_summ$features[1]`        | `r mia_me_asv_summ$features[1]`        |
| No. singleton ASVs    | `r mia_me_asv_raw_summ$features[2]`        | `r mia_me_asv_summ$features[2]`        |
| Avg ASVs per sample.  | `r round(mia_me_asv_raw_summ$features[3])` | `r round(mia_me_asv_summ$features[3])` |

We started off with `r nrow(me_asv_raw$tax_table)` ASVs and `r nrow(me_asv_raw$sample_table)` samples. Screening for `NA` kingdom assignment removed an additional `r nrow(me_asv_raw$tax_table) - nrow(me_asv_no_na$tax_table)` ASVs. Screening for Mitochondria and Chloroplast removed `r nrow(me_asv_no_na$tax_table) -  nrow(me_asv_no_cont$tax_table)` ASVs. After removing the negative controls there were `r nrow(me_asv_no_nc$tax_table)` ASVs and `r nrow(me_asv_no_nc$sample_table)`. After removing low-count samples, there were `r nrow(me_asv$tax_table)` ASVs and `r nrow(me_asv$sample_table)` samples remaining.

```{r}
#| echo: false
#| eval: false
objects()
gdata::keep(ds_metrics, sample_tab, sample_info, tax_tab, otu_tab, 
            me_asv_raw, me_asv_no_na, me_asv_no_cont, me_asv_nc, 
            me_asv_no_nc, me_asv_no_low, me_asv, nc_asvs, nc_check, 
            nc_remove, nc_remain, rem_nc_reads, rem_sam_reads, per_reads_rem, 
            per_reads_ret, ret_nc_reads, ret_sam_reads, mia_me_asv_raw, 
            mia_me_asv, mia_me_asv_raw_summ, mia_me_asv_summ, pipe_summary,
            sure = TRUE)
rm(list = ls(pattern = "tmp_"))
save.image("16S/LOTUS3/lotus3_processing.rdata")
```

